---
# YAML header ----
# qmd yaml chunk (note the 3 x '-' at the begining and end of the chunk)
# description: 
#  Use the yaml to:
#   - control document metadata-files: e.g., title, subtitle, description, abstract, date, bibliography.
#   - set document formating: e.g. document output type (html, docx, pdf, pptx), table of contents (toc), html behavior.
#       - For html see more here: https://quarto.org/docs/output-formats/html-basics.html
#       - For html see more here: https://quarto.org/docs/output-formats/pdf-basics.html
#       - For html see more here: https://quarto.org/docs/output-formats/ms-word.html
#   - html style can be handled by external style file (e.g. data_dev.css).
#   - If you want to force a line-break in the yaml use html '<br>'.
#   - Tabs (or spaces) and important in the yml (and main text). Indentation signals a sub option (or section).
# COMMENT = #

## Document metadata ----
title: "Scientific Services, BSPI Project report"
subtitle: "Species distribution models ― test runs"
author: "Paul Gordijn"
date: last-modified
date-format: "YYYY-MM-DD"
abstract: "Use this file as a project documentation template. <br> Insert here a sentence or two describing the project. List the 'umbrella' project that this report falls under where relevant. <br> "

## Document style ----
### html style ----
title-block-banner: true
format: 
    html:
        embed-resources: true
        smooth-scroll: true
        theme:
            - /home/paulg/.ekznw_quarto/style/_ekznw_theme.scss
        fontcolor: black
        toc: true
        toc-location: left
        toc-title: Summary
        toc-depth: 3
        number-sections: true

### pdf style and options ----
    pdf:
        toc: true     # appears as a PDF ToC page
        number-sections: true

### Word document stype and options ----
    docx:
        number-sections: true
css: /home/paulg/.ekznw_quarto/style/default_overrides.css

# Select bibliography ----
#  Cite with citation key like @cite_key or [@cite_key] for citations without,
# or within parentheses, respectively.
#  References will be printed in full at the end of the document.
bibliography: /home/paulg/refs/bspi.bib  # the source bib file can be downloaded from here:
                        # https://github.com/ekznw/refs

execute-dir: /home/paulg/dta_scp/
---

```{=html}
<!-- Quarto cheatsheet: https://rstudio.github.io/cheatsheets/quarto.pdf

In the Quarto main document wherein the write-up can be meshed with code. You can insert markdown comments that are not rendered into html, word etc using a 'markdown' block — note the beginning and end 3 x ` characters. This sequence of special characters in Quarto tell the renderer that this is special section, particularly, a comment section.   
See https://quarto.org/docs/authoring/callouts.html for more on callout blocks. There is a callout 'tip' box below. -->
```

<!-- ::: {.callout-tip title="NB!"}
The contents of this report should enable replication or further development of the project and associated data products.
::: -->

# Revision history

<!-- ::: {.callout-tip title="Tip"}
Remove contents for this project report template info below; then, for your project, add rows for major revisions or comments. Follow *version convention*: pre-release = v0.1 – v0.9; release \> v0.9.
::: -->

::: {#tbl-rev}
| Date | Person(s) | Note | Version |
|:----------------|:----------------|:---------------------|:----------------|
| 2026-02-09 | Adrian Armstrong & Paul Gordijn | Extract frog data and run rudimentary model. | v0.1 |

Project documentation revision history and version number.
:::

# Project summary

Work in progress.

## Purpose

<!-- ::: {.callout-tip title="What to write"}
A note on the 'why' of the project.
::: -->
Testing pipeline development for species Distribution Modelling.

# Main steps

<!-- ::: {.callout-tip title="What to write"}
List the main steps of the project. Keep it to the point—elaborate under 'Data processing' below.
::: -->

## First step (e.g. data prep)

## Other main processing steps

## ...

# Datasets used

<!-- ::: {.callout-tip title="What to write"}
Dataset citation(s) from BibTeX or other citation database.
::: -->

- @brun2025cba-a

# Data processing

<!-- ::: {.callout-tip title="What to write"}
Detailed elaboration of the 'main steps'. Should enable full reproduction of the project.
::: -->

## Processing environment

### R package management
```{r}
#| echo: true
#| message: false

# list, load R packages
pkgs <- c("ekznwr", "ENMeval", "kableExtra", "leaflet",
  "readxl", "sf", "terra", "usdm"
)

to_install <- pkgs[!pkgs %in% rownames(installed.packages())]

if (length(to_install)) {
  install.packages(to_install, dependencies = TRUE)
}
invisible(lapply(pkgs, library, character.only = TRUE))
```

### Directories and data

We'll be pulling data from the SCP data folder, and saving outputs in the working directory's data/outputs folder for now.
```{r}
datad <- file.path("~", "dta_scp")
workd <- normalizePath(".")

dir.create(file.path(workd, "tmp"), recursive = TRUE,
  showWarnings = FALSE
)
terraOptions(
  tempdir = file.path(workd, "tmp"),
  memfrac = 0.8,
  progress = 1
)

# add ekznwr options for archived data dir
options("extract_dir" = "~/.dta_cplan")
options("max_map_age" = 86400 / 4)
```

### Read in occurrence data
```{r}
#| eval: true
#| tbl-cap: Amphibian occurrence record summary (n = number of records) with first and lastest date of occureence reporting.
#| label: tbl-ampbib-summary
#| code-summary: Occurrence-data-summary
#| code-fold: true

occ <- dta_find(datad, search_key = "amphibian records") |>
  read_excel() |>
  setDT()

# filter out course spatial data (>250 m accuracy)
occ <- occ[SpatResol < 4]
occ$Date <- as.Date(occ$Date)

# summarize the taxonomic records
taxsum <- occ[,
  .(Species = unique(TaxonName), n = .N,
    firstocc = min(Date, na.rm = TRUE), lastocc = max(Date, na.rm = TRUE)
  ),
  by = TaxonID
][order(Species)]

# define species that will be modelled
spp <- c("Afrixalus spinifrons spinifrons",
  "Anhydrophryne ngongoniensis",
  "Breviceps bagginsi",
  "Breviceps carruthersi",
  "Breviceps passmorei",
  "Cacosternum rhythmum",
  "Hemisus guttatus",
  "Hyperolius pickersgilli",
  "Leptopelis xenodactylus"
)
taxsum <- taxsum[, to_model := fifelse(Species %chin% spp, "*", "")][,
  .(TaxonID, to_model, Species, n, firstocc, lastocc)
]

# format table for quarto report
kable(taxsum,
    col.names = c("id", "", "Species", "n", "First date", "Last date")
  ) |>
  kable_styling() |>
  column_spec(3, italic = TRUE)
```

# Plot the occurrence data

In this test round the species for which the tests are being run were restricted to:

- *Afrixalus spinifrons spinifrons*
- *Anhydrophryne ngongoniensis*
- *Breviceps bagginsi*
- *Breviceps carruthersi*
- *Breviceps passmorei*
- *Cacosternum rhythmum*
- *Hemisus guttatus*
- *Hyperolius pickersgilli*
- *Leptopelis xenodactylus*

```{r}
#| code-fold: true
#| fig-cap: Distribution of Amphibian occurrence records.
#| label: fig-frog-occ

occ <- occ[TaxonName %chin% spp]
occmap <- occ[, .(RecordID, TaxonName, DDS, DDE)]

occmap_sf <- st_as_sf(occmap, coords = c("DDE", "DDS"), crs = 4326)

species_list <- split(occmap_sf, occmap_sf$TaxonName)

m <- leaflet() |>
  addProviderTiles("CartoDB.Positron", group = "Light") |>
  addProviderTiles("Esri.WorldImagery", group = "Satellite") |>
  addProviderTiles("OpenTopoMap", group = "Terrain")

for (sp in names(species_list)) {
  df <- species_list[[sp]]   # this is already sf

  m <- addCircleMarkers(
    m,
    data = df,
    radius = 5,
    stroke = FALSE,
    fillOpacity = 0.7,
    group = sp,
    popup = ~paste0(
      "<b>Species:</b> ", TaxonName, "<br>",
      "<b>Record ID:</b> ", RecordID
    )
  )
}

m |>
  addLayersControl(
    baseGroups = c("Light", "Satellite", "Terrain"),
    overlayGroups = names(species_list),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  hideGroup(names(species_list)[-1])
```

## Model predictors (environmental data)

Make a table of available predictors.

swi, bioclim vars

```{mermaid}

```

```{r}
swi_fp <- dta_find(datad, "swi_s12_lo31.tif")
names(swi_fp) <- "swi"
bioc_fp <- dta_find(datad, "za-crop", dir_filter = "outputs")
names(bioc_fp) <- paste0("bioclim_", 1:19)
# get dem file path in zip archive
dem_arc <- dta_arc_extr( # extracts the tif from the ekznwr archive
  parent_dir = datad,
  arc_file = "srtmdem_wdd"
)
dem <- dta_fp( # obtains the file path
  extract_i = dem_arc, row_n = 9
)
names(dem) <- "dem"
env <- c(dem, swi_fp, bioc_fp)
```

Use the DEM as the template and resample other rasters to the template.
```{r}
template_name <- "dem"
template <- rast(env[template_name])
# check if the output already exists
if (!file.exists(file.path(workd, "data", paste0(template_name, ".tif")))) {
  template <- terra::project(
    template, "EPSG:32736", filename = file.path(workd, "data", "dem.tif"),
    overwrite = TRUE
  )
}
```

Resample rasters to lower(est) resolution of the mask.
```{r}
vars_to_trans <- env[!names(env) %in% template_name]
# first check if the files exist
output_name <- sapply(seq_along(vars_to_trans), function(i) {
  fn <- file.path(workd, "data", paste0(names(vars_to_trans[i]), ".tif"))
})
vars_to_trans <- vars_to_trans[!file.exists(output_name)]
lapply(seq_along(vars_to_trans), function(i) {
  project(
    rast(vars_to_trans[i]), # read in the raster
    template, # project to the template
    method = "near",
    filename = file.path(workd, "data", paste0(names(vars_to_trans[i]), ".tif")),
    overwrite = TRUE
  )
})
```

```{r}
#| eval: false

envars <- dta_find(file.path(workd, "data"), ".tif$")
envs <- rast(envars)
names(envs) <- gsub(".tif$", "", basename(envars))

# check for variance inflation between env variables
envs.vif <- usdm::vifstep(envs)
```

  - Include all env var for the model (standard model). Dev standard models for different taxon groups.
  - other vars: RH (and ~seasonality); Geology; Soils; Landform; Solar radiation; Wind; Veg; Land cover & transformation (most likely for post processing);  

# Products

::: {.callout-tip title="What to write"}
Notes on the final report and associated products.
:::

## Conclusion

::: {.callout-tip title="What to write"}
Concluding remarks.
:::

# Appendices

## Appendix A: metadata

::: {.callout-tip title="What to write"}
More on this to come...
:::

## Appendix B: supporting documentation

::: {.callout-tip title="What to write"}
List supporting documentation if it needs to be highlighted.
:::

## Appendix C: Project folder structure

```         
├── 📁 data         | Data folder
│   ├── 📁 arc      | Folder for archived data.
│   ├── 📁 dev      | Folder for datasets in development & temp files.
│   └── 📁 src      | Externally sourced data.
├── 📁 images       | Miscellaneous images.
│   └── 🖼️ graph01.png | Misc figs etc
├── 📁 .ekznw_quarto                  | Quarto document style
│   ├── 🎨 _ekznw_theme.scss          | Html theme.
│   ├── 🎨 default_overrides.css
├── 🌐 data_dev.html | Rendered html output
└── 📄 data_dev.qmd  | Raw Quarto document
```

## Appendix D: Processing environment

### R environment

```{r}
#| eval: true
sessionInfo()
```

# References (example references)